A formal treatment of bidirectional typing

### 2023-10-13 tex/ESOP24/BiSig.tex
[ tex ] Update Example section
[ tex ] Rewrite Formalisation

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mostly on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in an abstract syntax tree given by programmers.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
\subfile{S7-formalisation}
\input{S8-future}
\appendix
\input{appendix}

@@

\IfFileExists{reference}{
\bibliography{reference}
}{\bibliography{ref}}

>>>>

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
\subfile{S7-formalisation}
\input{S8-future}

@@

\IfFileExists{reference.bib}{
\bibliography{reference}
}{\bibliography{ref}}

\appendix
\input{appendix}

### 2023-10-13 tex/ESOP24/S5-synthesis.tex
[ tex ] Rewrite Formalisation

\begin{remark}
  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules}, not for derivations.
  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems even in the absence of a generic definition of bidirectional type systems.
\end{remark}

It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct manually by definition or automatically by the following lemma:
\begin{lemma}
  It is decidable if an operation $\biop$ is mode-correct or not.
\end{lemma}

>>>>

\begin{remark}
  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules}, not for derivations.
  Thus, this property cannot be formulated without treating rules as some mathematical object such as those general definitions in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems in the absence of a general definition.
\end{remark}

It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
\begin{lemma}\label{lem:decidability-mode-correctness}
  It is decidable whether an operation $\biop$ is mode-correct.
\end{lemma}

### 2023-10-13 tex/ESOP24/S6-example.tex
[ tex ] Rewrite Formalisation

\section{Examples}\label{sec:example}
In this section, we use two more examples to exhibit the applicability of our approach.
This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language.

@@

At first glance, accommodating this form of application may seem impossible, given that the number of arguments a construct can accept has to be fixed in a signature.
Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.

To illustrate this, let us consider the following rule:
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
We use the same type signature $\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
for each $n : \N$, so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.
\subsection{Computational calculi}\label{subsec:PCF}
The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can reduce the effort significantly.
For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.

@@

The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
However, if we implement a proof-relevant type synthesiser following \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.

>>>>

\section{Examples}\label{sec:example}
Now we use two more examples to exhibit the applicability of our approach.
This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language to justify the practical needs for a general development.

@@

At first glance, accommodating this form of application may seem impossible, given that the number of arguments for a construct is finite and has to be fixed.
Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
as a rule schema parametrised by $n$.
Then, we use the same type signature~$\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$, thereby exhibiting the necessity of having an arbitrary set for operation symbols.

\subsection{Computational calculi}\label{subsec:PCF}
The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can still reduce the effort significantly.
For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.

@@

The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
However, if we implement a proof-relevant type synthesiser in the style of \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.
On the other hand, similar to a parser generator, a type-synthesiser generator only needs a language specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
We will see in the next section how the actual formalisation is used.
